---
title: "Class 19: Pertussis and the CMI-PB project"
author: "Lena (A16420052)"
format: pdf
---

## Investigating pertussis cases by year

Pertussis is a severe lung infection also known as whopping cough. 

We will begin by investigating the number of Pertussis cases per year in the US.

This data is available on the CDC website [here](https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

```{r}
#| echo=FALSE
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```

Lets have a wee look at this data.frame

```{r}
head(cdc)
```

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.I want a nice plot of the number of cases per year. 

```{r}
library(ggplot2)

ggplot(cdc, aes(Year, Cases)) + geom_point() + geom_line()

```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
library(ggplot2)

ggplot(cdc, aes(Year, Cases)) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept=1946, linetype="dashed", col="blue") + 
  geom_vline(xintercept=1996, linetype="dashed", col="red") + 
  geom_vline(xintercept=2020, linetype="dashed", col="purple")
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, cases rose slightly and had a small peak in 2012. A possible explanation could be the anti-vax movement, bacterial immunity. 

## Exploring CMI-PB data

This is the goals of the CMI-PB project: http://www.cmi-pb.org/

The CBI-PB project makes its data avaiable via "API-endpoint" that returns JSON format 

We will use the **jsonlite** package to access this data. The main function in this package is called `read_json()`


```{r}
library(jsonlite)

subjects <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE)
titer <- read_json("https://www.cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE)
```

Have a wee peek at these new objects

```{r}
head(subjects)
head(specimen)
head(titer)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subjects$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subjects$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subjects$race, subjects$biological_sex)
```

## Working with dates

Dates can really suck to work with and do math. Lubricate package makes this easier. It is a part of the **tidyverse** that includes dplyr, ggplot2, etc. 

```{r}
library("tidyverse")
```

```{r}
time_length(today()-mdy("07-25-2002"), "years")

```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subjects$age <- ymd(subjects$date_of_boost) - ymd(subjects$year_of_birth)

subjects$age_years <- time_length(subjects$age, "years")

head(subjects)
```


```{r}
library(dplyr)
ap <- subjects %>% 
  filter(infancy_vac=="aP") 
round(summary(time_length(ap$age, "years")))

wp <- subjects %>% 
  filter(infancy_vac=="wP") 
round(summary(time_length(wp$age, "years")))


```

```{r}
# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value

```

It is significantly different. 

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subjects$date_of_boost) - ymd(subjects$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


```{r}
library(dplyr)
ap <- subjects %>% 
  filter(infancy_vac=="aP") 
round(summary(time_length(ap$age, "years")))

wp <- subjects %>% 
  filter(infancy_vac=="wP") 
round(summary(time_length(wp$age, "years")))


```
> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subjects) +
  aes(age_years,
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

Merge or join tables 

Link subjects and specimen using dplyr `*_join()` function

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subjects)
head(meta)
```

Antibody measurements in the blood

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
head(abdata)

```
How many isotypes are we measuring for all these individuals?

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

Lets focus on one of these IgG

```{r}
igg <- abdata %>% filter(isotype=="IgG")
head(igg)
```

Boxplot of MFI _normalized vs antigen 

```{r}
ggplot(igg, aes(MFI_normalised, antigen)) + geom_boxplot()
```

```{r}
ggplot(igg, aes(MFI_normalised, antigen, col=infancy_vac)) + geom_boxplot()
```

Focus in on IgG to the Pertussis Toxin (PT) antigen in teh 2021 dataset

```{r}
igg.pt <- igg %>% filter(antigen=="PT", dataset=="2021_dataset")
```

```{r}
ggplot(igg.pt, aes(planned_day_relative_to_boost, MFI_normalised, col=infancy_vac, group=subject_id)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=0, linetype="dashed", col="black") +
  geom_vline(xintercept=14, linetype="dashed", col="black")
```

> Q18. Does this trend look similar for the 2020 dataset?

14 days for it to peak. wP higher than aP?
